---
title: "HW 4"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter your name and EID here: Joseph Hendrix \| jlh7459

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

------------------------------------------------------------------------

### Question 1: (2 pts)

All subsequent code will be done using `dplyr`, so we need to load this package. We also want to look at the `penguins` dataset which is inside the `palmerpenguins` package:

```{r message=FALSE}
# Call dplyr and ggplot2 packages within tidyverse
library(tidyverse)

# Paste and run the following uncommented code into your console: 
# install.packages("palmerpenguins")

# Save the data as a dataframe
penguins <- as.data.frame(palmerpenguins::penguins)
```

Using a `dplyr` function, pick all the rows/observaions in the `penguins` dataset from the year 2007 and save the result as a new object called `penguins_2007`. Compare the number of observations/rows in the original `penguins` dataset with your new `penguins_2007` dataset.

```{r}
# filtering penguins data set
penguins_2007 <- penguins %>% filter(year == 2007)
penguins
penguins_2007

```

**The full penguins dataset has 344 observations, 110 of which are from 2007**

------------------------------------------------------------------------

### Question 2: (2 pts)

Using `dplyr` functions on `penguins_2007`, report the number of observations for each species-island combination (note that you'll need to `group_by`). Which species appears on all three islands?

```{r}
# pulling species-island count
penguins_2007 %>%
  group_by(species, island) %>%
  summarize(n())
  
```

**Adelie penguins appear on all three islands.**

------------------------------------------------------------------------

### Question 3: (2 pts)

Using `dplyr` functions on `penguins_2007`, create a new variable that contains the ratio of `bill_length_mm` to `bill_depth_mm` (call it `bill_ratio`). *Once you checked that your variable is created correctly*, overwrite `penguins_2007` so it contains this new variable.

```{r}
# your code goes below (make sure to edit comment)
penguins_2007$bill_length_mm / penguins_2007$bill_depth_mm    # verifying

penguins_2007 %>%
  mutate(bill_ratio = (bill_length_mm / bill_depth_mm))    # adding to dataset
```

Are there any cases in the `penguins_2007` dataset for which the `bill_ratio` exceeds 3.5? If so, for which species of penguins is this true?

```{r}
# select observations where bill ratio is greater than 3.5
penguins_2007 %>%
  mutate(bill_ratio = (bill_length_mm / bill_depth_mm)) %>%
  filter(bill_ratio > 3.5)
```

**The Gentoo penguin has a bill ratio greater than 3.5**

------------------------------------------------------------------------

### Question 4: (2 pts)

Using `dplyr` functions on `penguins_2007`, find the three penguins with the smallest bill ratio for *each species*. Only display the information about `species`, `sex`, and `bill_ratio`. Does the same sex has the smallest bill ratio across species?

```{r}
# your code goes below (make sure to edit comment)
penguins_2007 %>%
  mutate(bill_ratio = (bill_length_mm / bill_depth_mm), na.rm=T) %>%    # create bill_ratio
  select(species, sex, bill_ratio) %>%    # select relevant info
  group_by(species, sex) %>%
  summarize(min(bill_ratio, na.rm=T))    # select minimum bill ratio values
```

**No! Males have smaller bill ratios among Adelie and Gentoo penguins but female Chinstraps have smaller bill ratios than male Chinstraps**

------------------------------------------------------------------------

### Question 5: (2 pts)

Using `dplyr` functions on `penguins_2007`, calculate the mean and standard deviation of `bill_ratio` for each species. Drop NAs from `bill_ratio` for these computations (e.g., using the argument `na.rm = T`) so you have values for each species. Which species has the greatest mean `bill_ratio`?

```{r}
penguins_2007 %>%
  mutate(bill_ratio = bill_length_mm / bill_depth_mm, na.rm = T) %>%    # create bill_ratio obs
  group_by(species) %>%
  summarize(mean(bill_ratio, na.rm=T), sd(bill_ratio, na.rm=T))    # calculate mean and sd of bill ratio, excluding NA values
```

**Gentoos have the greatest average bill ratio.**

------------------------------------------------------------------------

### Question 6: (2 pts)

Using `dplyr` functions on `penguins_2007`, remove missing values for `sex`. Pipe a `ggplot` to create a single plot showing the distribution of `body_mass_g` colored by male and female penguins, faceted by species (use the function `facet_wrap()` with the option `nrow =` to give each species its own row). Which species shows the least sexual dimorphism (i.e., the greatest overlap of male/female size distributions)?

```{r}
penguins_2007 %>%
  select(body_mass_g, sex, species) %>%    # select relevant info
  filter(!is.na(sex)) %>%    # exclude obs where sex is NA
  ggplot(aes(body_mass_g, sex, fill = sex)) +    #color bars according to sex
  geom_bar(stat="identity") +
  facet_wrap(~species, nrow = 3)
```

**Chinstrap penguins exhibit the least sexual dimorphism; their average body mass is very similar.**

------------------------------------------------------------------------

### Question 7: (2 pts)

Pipe a `ggplot` to `penguins_2007` to create a scatterplot of `body_mass_g` (y-axis) against `bill_ratio` (x-axis). Does it look like there is a relationship between the bill ratio and the body mass? *Note: you might see a Warning message.* What does this message refer to?\*

```{r}
penguins_2007 %>%
  mutate(bill_ratio = bill_length_mm / bill_depth_mm, na.rm=T) %>%    # create bill ratio obs
  select(body_mass_g, bill_ratio) %>%    # select relevant obs
  filter(!is.na(bill_ratio) & !is.na(body_mass_g)) %>%    # exclude missing values for either variable
  ggplot(aes(bill_ratio, body_mass_g)) +
  geom_point()
```

**There appears to be a positive linear relationship between body mass and bill ratio. I didn't get a warning message, but I imagine it would be due to missing values in either variable.**

------------------------------------------------------------------------

### Question 8: (2 pts)

What if we separate each species? Duplicate the plot from the previous question and add a regression trend line with `geom_smooth(method = "lm")`. Color the points and the regression lines by species. Does the relationship between the bill ratio and the body mass changes within each species?

```{r}
penguins_2007 %>%
  mutate(bill_ratio = bill_length_mm / bill_depth_mm, na.rm=T) %>%    # create bill ratio obs
  select(body_mass_g, bill_ratio, species) %>%    # select relevant obs
  filter(!is.na(bill_ratio) & !is.na(body_mass_g)) %>%    # exclude missing values for either variable
  ggplot(aes(bill_ratio, body_mass_g, color=species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

**The relationship changes! It appears that Adelie bill ratio increases as body mass decreases.**

------------------------------------------------------------------------

### Question 9: (2 pts)

Finally, let's make a plot using the original `penguins` dataset (not just the 2007 data). Forewarning: This will be very busy plot!

Map `body_mass_g` to the y-axis, `flipper_length_mm` to the x-axis, `species` to color, and `island` to shape. Using `facet_wrap()`, facet the plots by `year`. Find a way to clean up the x-axis labels (e.g., reduce the amount of tick marks) using `scale_x_continuous()`. Does there appear to be a relationship between body mass and flipper length overall? Is there a relationship within each species? What happens to the distribution of flipper lengths for species over time?

```{r}
penguins %>%
  select(body_mass_g, flipper_length_mm, species, island, year) %>%   # select data
  filter(!is.na(body_mass_g) & !is.na(flipper_length_mm)) %>%    # remove missing values
  filter(!is.na(year)) %>%    # more filtering woohoo
  ggplot(aes(flipper_length_mm, body_mass_g, color=species, shape=island)) +
  geom_point() +
  scale_x_continuous(breaks = c(170, 190, 210, 230)) +    # set x-axis breaks
  facet_wrap(~year, nrow = 3)    # distinct year graphs
```

**There appears to be a positive linear relationship between body mass and flipper length throughout the dataset; flipper length increases as body mass increases. This holds for all species and across time. The flipper length distribution shifted to the right across time; the penguins' flipper length grew with age.**

------------------------------------------------------------------------

### Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```
