---
title: "HW 6"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE,  
                      warning = FALSE, message = FALSE, 
                      fig.align = "center",
                      R.options = list(max.print=100))

# Edit the file starting below
```

### Enter your name and EID here: Joseph Hendrix | jlh7459

**You will submit this homework assignment as a pdf file on Gradescope.**

*For all questions, include the R commands/functions that you used to find your answer (show R chunk). Answers without supporting code will not receive credit. Write full sentences to describe your findings.*

We will use the packages `tidyverse` and  `plotROC` for this assignment.

```{r, message = FALSE}
# Load packages
library(tidyverse)
library(plotROC)
```

------------------------------------------------------------------------

## Question 1: (4 pts)

We will use the `pokemon` dataset for this assignment:

```{r}
# Upload data from GitHub
pokemon <- read_csv("https://raw.githubusercontent.com/laylaguyot/datasets/main//pokemon.csv")

# Take a look 
head(pokemon)
```

Recode the variable `Legendary`, taking a value of 1 if a pokemon is legendary and a value of 0 if it is not. Save the resulting data as `my_pokemon`.

```{r}
# recode Legendary variable
my_pokemon = pokemon %>%
  mutate(Legendary = ifelse(Legendary == TRUE, 1, 0))

my_pokemon
 
```

Let's visualize how the features of `Attack` and `HP` impact the legendary status. First, visualize the distribution of `Attack` for legendary pokemons vs those that are not. Also visualize the distribution of `HP` for these two groups. *Note: consider the binary variable as a factor for your `ggplot` using `as.factor()`.* Comment with what you see in these visualizations.

```{r}
# recode legendary as factor and plot against attack
my_pokemon %>%
  mutate(fac = as.factor(Legendary)) %>%
  select(Attack, fac) %>%
  ggplot(aes(x = Attack, y = fac)) +
  geom_boxplot() +
  labs(title = "Attack Stats for Legendaries and Non-Legendaries", y = "Legendary Status")

# recode legendary as factor and plot against HP
my_pokemon %>%
  mutate(fac = as.factor(Legendary)) %>%
  select(HP, fac) %>%
  ggplot(aes(x = HP, y = fac)) +
  geom_boxplot() +
  labs(title = "HP Stats for Legendaries and Non-Legendaries", y = "Legendary Status")
```

**Legendary pokemon tend to have higher average Attack and HP stats; non-legendaries have a larger variance for both variables.**

Then visualize the linear relationship between `Attack` and `HP` (hit points) for each legendary status. *Hint: color the regression lines.* Do `Attack` and `HP` seem to predict Legendary status? Comment with what you see in this visualization.

```{r}
# plot linear relationship between attack and hp, legendaries
my_pokemon %>%
  filter(Legendary ==  1) %>%
  select(Attack, HP) %>%
  ggplot(aes(Attack, HP)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = 'red') +
  labs(title = "HP vs Attack for Legendary Pokemon")

# plot same for non-legendaries
my_pokemon %>%
  filter(Legendary ==  0) %>%
  select(Attack, HP) %>%
  ggplot(aes(Attack, HP)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = 'blue') +
  labs(title = "HP vs Attack for Non-Legendary Pokemon")

 
```

**There seems to be a linear relationship between HP and Attack.**

------------------------------------------------------------------------

## Question 2: (2 pt)

Let's predict `Legendary` status using a linear regression model with `Attack` and `HP` in `my_pokemon`. Fit this model, call it `pokemon_lin`, and write its equation.

```{r}
# fit linear model to attack and hp to predict if a pokemon is a legendary
pokemon_lin = lm(Legendary ~ Attack + HP, data = my_pokemon)
summary(pokemon_lin)
```

**Legendary = -0.22 + 0.003Attack + 0.002HP**

------------------------------------------------------------------------

## Question 3: (3 pts)

Choose a pokemon whose name starts with the same letter as yours. Take a look at its stats and, using the equation of your model from the previous question, predict the legendary status of this pokemon, "by hand":

```{r}
# select a pokemon that starts with 'j', prolly jolteon he was my favorite eeveelution as a kid
# nvm joltik is cooler bs it evolves into galvantula
my_pokemon %>%
  filter(str_detect(Name, "^J"))

# predict by hand
joltik = my_pokemon %>%
  filter(Name == "Joltik")

-0.22 + (joltik$Attack * 0.003) + (joltik$HP * 0.002)
# Legendary = 0.021 = not a legendary
```

Check your answer by using `predict()` with the argument `newdata =`:

```{r}
# predict joltik's legendary status using predict()
my_pokemon %>%
  mutate(predictions = predict(pokemon_lin)) %>%
  filter(Name == "Joltik")
```

Was your pokemon predicted to be legendary? Why or why not? Does it match the reality?

**Since the predicted value is less than one, Joltik is predicted to be a non-legendary pokemon. This is true!**

------------------------------------------------------------------------

## Question 4: (2 pts)

We can measure how far off our predictions are from reality with residuals. Use `resid()` to find the residuals of each pokemon in the dataset then find the sum of all residuals. Why does it make sense? 

```{r}
# calculate residuals
my_pokemon %>%
  mutate(predictions = predict(pokemon_lin)) %>%
  mutate(residuals = resid(pokemon_lin)) %>%
  summarize(sum(residuals))
```

**The sum of residuals is pretty small since we expect the positive residuals to cancel out the negative ones**

------------------------------------------------------------------------

## Question 5: (2 pts)

A logistic regression would be more appropriate to predict `Legendary` status since it can only take two values. Fit this new model with `Attack` and `HP`, call it `pokemon_log`, and write its equation. *Hint: the logit form is given by the R output.*

```{r}
#fit logistic model to data
pokemon_log = glm(Legendary ~ Attack + HP, data = my_pokemon)
pokemon_log
```

**Equation: -0.220 + (0.002)Attack + (0.002)HP**

------------------------------------------------------------------------

## Question 6: (2 pts)

According to this new model, is the pokemon you chose in question 3 predicted to be legendary? Why or why not? *Hint: you can use predict() with the arguments `newdata =` and `type = "response"`.*

```{r}
# predict if joltik is a legendary using logistic regression
my_pokemon %>%
  mutate(predictions = predict(pokemon_log)) %>%
  filter(Name == "Joltik")
```

**Again, since Joltik has a value less than 1 he's predicted to be a non-legendary**

------------------------------------------------------------------------

## Question 7: (3 pts)

Let's compare the performance of these two models using ROC curves. On the same plot, represent the ROC curve for predicting `Legendary` status based on the predictions from the linear regressionin blue and another ROC curve based on the predictions from the logistic regression in orange.

```{r}
# plot ROC curves
ROC = my_pokemon %>%
  mutate(pLin = predict(pokemon_lin)) %>%
  mutate(pLog = predict(pokemon_log)) %>%
  ggplot() +
  geom_roc(aes(d = Legendary, m = pLin, color = 'blue')) +    # linear
  geom_roc(aes(d = Legendary, m = pLog, color = "orange"))    # logistic

ROC
```

How do these two models compare?

**It seems both models produce the same ROC curve!**

------------------------------------------------------------------------

## Formatting: (2 pts)

Comment your code, write full sentences, and knit your file!

------------------------------------------------------------------------

```{r, echo=F}
## DO NOT DELETE THIS BLOCK!
Sys.info()
```